### 2018-07-06 submission/manuscript.tex
add result updates

\textbf{Background.} A typical 16S rRNA gene sequencing workflow can be
divided into preservation, extraction, amplification, and sequencing
steps. At each of these stages error can be introduced that will change
the underlying bacterial community composition results. In this study we
focus on the amplification step's contribution to this overall error. To
accomplish this we assessed 16S rRNA gene sequencing results in human
fecal and mock community samples after using different high fidelity
(HiFi) DNA polymerases and number of amplification cycles.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. We first
assessed GC dependent differences, error rate, sequence error
prevalence, chimera prevalence, and correlation between chimera
prevalence and number of Operational Taxonomic Units (OTUs) by
polymerase and number of cycles. Next, differences in the number of OTUs
and taxa was examined based on the polymerase and number of cycles used.
Additionally, differences in the bacterial community composition by the
Bray-Curtis index also was assessed based on polymerase and number of
cycles. Finally, Random Forest models were created to test whether the
bacterial community was better at classifying polymerases, number of
cycles, or individual donor. We also assessed whether the most important
taxa in the polymerase and number of cycle Random Forest models also
were the most important in the model for individual donors.

\textbf{Results.} In addition to the total counts for specific taxa
varying by polymerase, we found noticeable differences in relative
abundance based on high and low GC content (P-value \(\leqslant\) 0.04).
Chimera prevalence in mock communities varied by polymerase with
differences being most notable at 35 cycles (Kappa = 5.71\% (median)
versus Platinum = 26.62\%) and this variation persisted after chimera
removal using VSEARCH. We also observed positive correlations between
chimera prevalence and the number of OTUs with Platinum having the
highest (R\textsuperscript{2} = 0.974) and Kappa having the worst
(R\textsuperscript{2} = 0.478). When analyzing mock community samples
the variation in the number of OTUs detected by the polymerases was
observable as early as 20 cycles (P-value = 0.002). There also was a
large range in the number of OTUs amplified by the polymerases at 35
cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73). When analyzing
fecal samples we observed that the range in the number of OTUs detected
was not consistent between HiFi DNA polymerases (eg. at 35 cycles
Accuprime = 84 - 106 (min - max) versus Phusion = 84 - 136).
Additionally, the median number of OTUs varied by HiFi DNA polymerase
used at 35 cycles (P-value \textless{} 0.0001). Random Forest models
were most successful at classifying individual donor samples rather than
polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
Additionally, the most important OTUs in the polymerase and number of
cycle models were not the most important in the individual donor sample
model.

@@

\textbf{\emph{Observed relative abundance differs by GC content.}} There
was a significant difference in relative abundance between high/low GC
content based on either the V4 16S rRNA gene region or the whole genome
{[}Figure 1 \& Table S1{]}.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used.}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
majority of the differences across the number of cycles was between
Kappa and the other polymerases {[}Figure 4 \& Table S4{]}. The total
sequences with at least one error was also polymerase dependent, with
the majority of differences being between Kappa or Accuprime and the
other polymerases {[}Table S3 \& S4{]}. These differences in error rates
were not due to polymerase dependent differences in base substitution
rate {[}Figure S1{]}. Collectively, the results suggest that sequence
error is dependent on polymerase, persists across the number of cycles
used, and are not due to any bias towards a specific base substitution.

\textbf{\emph{Prevalence of chimeric sequences are polymerase dependent
and correlate with the number of OTUs:}} Based on the previous results,
we examined whether chimeras also were dependent on polymerase and
whether this could affect the number of OTUs. We observed significant
differences in the chimera prevalence based on polymerase across the
number of cycles used (P-value \textless{} 0.05) {[}Table S3{]}.
Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S4{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 5A \& 5B{]}. The number of cycles used clearly
increased chimera prevalence for all polymerases but the rate of
increase differed {[}Figure 5A \& 5B{]}. Additionally, there was a
plateau in the total percent of chimeras that were removed that was
similar for all polymerases {[}Figure 5C{]}. A positive correlation was
observed between chimeric sequences and the number of OTUs for all
polymerases {[}Figure 6{]}. This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
This data suggests that chimera prevalence depends on polymerase, is
made worse by increasing the number of cycles, and confirms that the
number of OTUs is dependent on the prevalence of these chimeric
sequences.

\textbf{\emph{The number of OTUs generated are dependent on polymerase
used:}} Differences in the range of the number of OTUs detected for
fecal samples is dependent on the polymerase used (e.g.~Accuprime at 35
cycles = 84 - 106 versus Phusion at 35 cycles = 84 - 136) {[}Figure
1{]}. Additionally, there is a trend for lower number of cycles (15-20)
to result in a reduced range in the number of OTUs versus higher number
of cycles (25, 30, and 35) for all polymerases (e.g.~Phusion at 15
cycles = 10 - 19 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}.
There is an overall difference in the number of OTUs detected within
fecal samples between polymerases at 35 cycles (F-stat \textgreater{}
16.35, P-value = 9.7e-05) {[}Table S1{]}, but a Tukey post-hoc test
failed to identify which specific polymerase groups were different
(P-value \textgreater{} 0.05) {[}Table S2{]}. The polymerase dependent
difference in the range of the number of OTUs also was observed in the
mock community samples {[}Figure 2{]}. The closest the polymerases came
to the total of 8 OTUs created by the mock reference 16S sequences was
at 25 and 30 cycles {[}Figure 2{]}. Regardless of if fecal or mock
communities were used, the same polymerases generated high and low
number of OTUs and this was consistent across the number of cycles used
{[}Figure 1-2 \& Table S1-S2{]}. In contrast to the results obtained
with fecal samples, differences between polymerases for the number of
OTUs created were observed as early as 20 cycles in the mock community
(F-stat = 15.82, P-value = 0.002) {[}Table S1{]}. Using a Tukey post-hoc
test, the majority of differences for the number of OTUs detected in the
mock community was largely due to Kappa and Platinum versus the other
polymerases across the number of cycles used {[}Table S2{]}. Based on
these observations in fecal and mock communities, it is clear that using
different polymerases result in a different total number of OTUs within
a sample.

\textbf{\emph{The bacterial communities generated by the polymerases
were similar but varied by number of cycles:}} Within each respective
5-cycle increment comparison there was no difference in Bray-Curtis
index between the polymerases (P-value \textgreater{} 0.05) {[}Figure
3{]}. Using PERMANOVA to test for community differences based on any of
the number of cycles within polymerases, only Phusion had cycle
dependent differences (P-value = 0.03). For fecal samples, Phusion was
one of two polymerases that had enough sequences to be rarefied to 1000
sequences at 15 cycles. There was an overall decrease in Bray-Curtis
index by cycle comparison group (e.g.~the 15 cycle versus 20 cycle group
compared to the 30 cycle versus 35 cycle group) (P-value \textless{}
0.01) {[}Figure 3A{]}. Using a Dunn's post-hoc test the 15 cycle versus
20 cycle and 20 cycle versus 25 cycle comparison groups had a higher
Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
(P-value \textless{} 0.05). Similar trends were observed in the mock
community but none were significant (P-value \textgreater{} 0.05)
{[}Figure 3B{]}. Overall, these data suggest that the number of cycles
can change the bacterial community independent of the polymerase used to
generate the sequences.

>>>>

\textbf{Background.} A typical 16S rRNA gene sequencing workflow can be
divided into preservation, extraction, amplification, and sequencing
steps. At each of these stages error can be introduced that will change
the underlying bacterial community composition results. In this study,
we focus on the amplification step's contribution to this overall error.
To accomplish this we assessed 16S rRNA gene sequencing results in human
fecal and mock community samples after using different high fidelity
(HiFi) DNA polymerases and number of amplification cycles.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. We first
assessed GC dependent differences, error rate, sequence error
prevalence, chimera prevalence, and correlation between chimera
prevalence and number of Operational Taxonomic Units (OTUs) by
polymerase and number of cycles. Next, differences in the number of OTUs
was examined based on the polymerase and number of cycles used.
Additionally, differences in the bacterial community composition by the
Bray-Curtis index also was assessed based on polymerase and number of
cycles. Based on these findings individual taxa differences based on
polymerase and number of cycles was investigated. Finally, Random Forest
models were created to test whether the bacterial community was better
at classifying polymerases, number of cycles, or individual donor. We
also assessed whether the most important taxa in the polymerase and
number of cycle Random Forest models were also important in the model
for individual donors.

\textbf{Results.} Predictably, we found noticeable differences in
relative abundance based on high and low GC content (P-value
\(\leqslant\) 0.04). Chimera prevalence in mock communities varied by
polymerase with differences being most notable at 35 cycles (Kappa =
5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
after chimera removal using VSEARCH. We also observed positive
correlations between chimera prevalence and the number of OTUs with
Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
having the worst (R\textsuperscript{2} = 0.478). When analyzing mock
community samples the variation in the number of OTUs detected by the
polymerases was observable as early as 20 cycles (P-value = 0.002).
There also was a large range in the number of OTUs amplified by the
polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
When analyzing fecal samples we observed similar differences in the
number of OTUs. However, the median number of OTUs only varied by HiFi
DNA polymerase used at 35 cycles and not at 20 cycles like the mock
community (P-value \textless{} 0.0001). Random Forest models were most
successful at classifying individual donor samples rather than
polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
Additionally, the most important OTUs in the polymerase and number of
cycle models were not the most important in the individual donor sample
model.

@@

\textbf{\emph{Mock community relative abundance differs by GC content.}}
There was a significant difference in relative abundance between
high/low GC content based on either the V4 16S rRNA gene region or the
whole genome {[}Figure 1 \& Table S1{]}. However, using only the GC
content of the V4 region resulted in less differences than when using
the whole genome {[}Table S1{]}. The differences between high and low GC
content groups were significant as early as 20 cycles in all polymerases
except Q5 {[}Table S1{]}. The highest (\emph{Staphylococcus}) and lowest
(\emph{Pseudomonas}) GC content bacteria were the most divergent from
the expected relative abundance of 0.12 {[}Figure 1{]}. These
observations suggest that the high fidelity DNA polymerases tested do
not prevent GC-based amplification error.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used.}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 2A \& Table S2{]}.
Thus, the majority of the differences observed across the number of
cycles was between Kappa and other polymerases {[}Figure 2A \& Table
S3{]}. The total sequences that contained at least one error was also
polymerase dependent, with the majority of differences being between
Kappa or Accuprime and the other polymerases {[}Table S2 \& S3{]}. These
differences in error rates were not due to polymerase dependent
differences in base substitution rate {[}Figure S1{]}. Collectively, the
results suggest that sequence error is dependent on polymerase, persists
across the number of cycles used, and are not due to any bias towards a
specific base substitution.

\textbf{\emph{Prevalence of chimeric sequences are polymerase dependent
and correlate with the number of OTUs.}} Based on the previous results,
we examined whether chimeras also were dependent on polymerase and
whether this could affect the total number of OTUs. We observed
significant differences in the chimera prevalence based on polymerase
across the number of cycles used (P-value \textless{} 0.05) {[}Table
S2{]}. Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
increased chimera prevalence for all polymerases but the rate of
increase differed {[}Figure 2B \& 2C{]}. Additionally, there was a
plateau in the total percent of chimeras that were removed that was
similar for all polymerases {[}Figure 2D{]}. A positive correlation was
observed between chimeric sequences and the number of OTUs for all
polymerases {[}Figure 2E{]}. This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
This data suggests that chimera prevalence depends on polymerase, is
made worse by increasing the number of cycles, and confirms that the
number of OTUs is dependent on the prevalence of these chimeric
sequences.

\textbf{\emph{The number of OTUs generated are dependent on polymerase
used.}} In addition to the sequence error and chimera differences
between polymerases, we also investigated whether there were polymerase
or number of cycle dependent effects on the observed number of OTUs.
There was a polymerase dependent difference in the range of the number
of OTUs in the mock community samples {[}Figure 3A{]}. The closest any
of the polymerases came to a total of 8 OTUs, created by running the
mock reference 16S sequences through mothur processing, was at 25 and 30
cycles {[}Figure 3A{]}. Differences between polymerases for the number
of OTUs created were observed as early as 20 cycles in the mock
community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
Tukey post-hoc test, the majority of differences for the number of OTUs
detected in the mock community was largely due to Kappa and Platinum
versus the other polymerases across the number of cycles used {[}Table
S5{]}. Similarily, differences in the range of the number of OTUs
detected for fecal samples was dependent on the polymerase used
(e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion at 35 cycles = 84
- 136) {[}Figure 3B{]}. There was also a trend for lower number of
cycles (15-20) to result in a reduced range in the number of OTUs versus
higher number of cycles (25, 30, and 35) for all polymerases
(e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35 cycles = 84 -
136) {[}Figure 3B{]}. Unlike mock community samples, an overall
difference in the number of OTUs was only detected between polymerases
at 35 cycles in fecal samples (F-stat \textgreater{} 16.35, P-value =
9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to identify
any differences between specific polymerase groups (P-value
\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
noticeably larger difference between polymerases in the number of OTUs
than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
consistent across the number of amplification cycles tested {[}Figure
3{]}. Based on these observations, it is clear that larger differences
in total number of OTUs based on polymerases are found in mock versus
fecal communities.

\textbf{\emph{The bacterial communities generated by the polymerases
were similar but varied by number of cycles.}} Next, we investigated the
effect polymerases and number of cycles had on the bacterial community
composition in more detail by using the Bray-Curtis index. We observed
no difference in the bacterial community composition between any measure
tested when using the mock community (P-value \textgreater{} 0.05)
{[}Figure 4A{]}. With respect to fecal samples, within each respective
5-cycle increment comparison there was no difference in Bray-Curtis
index between the polymerases (P-value \textgreater{} 0.05) {[}Figure
4B{]}. Using PERMANOVA to test for community differences based on any of
the number of cycles within polymerases, only Phusion had cycle
dependent differences (P-value = 0.03). Additionally, Phusion was one of
two polymerases that had enough sequences to be rarefied to 1000
sequences at 15 cycles. There was an overall decrease in Bray-Curtis
index by cycle comparison group (e.g.~the 15 cycle versus 20 cycle group
compared to the 30 cycle versus 35 cycle group) (P-value \textless{}
0.01) {[}Figure 4B{]}. Using a Dunn's post-hoc test the 15 cycle versus
20 cycle and 20 cycle versus 25 cycle comparison groups had a higher
Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
(P-value \textless{} 0.05). Overall, these data suggest that the number
of cycles can change the bacterial community independent of the
polymerase used to generate the sequences.

\textbf{\emph{Polymerase dependent differences in taxa are more
pronounced in mock versus fecal samples.}} Finding minimal differences
in overall bacterial community composision, we next examined if specific
taxa vary depending on polymerase and number of cycles. Further
investigation of mock community samples found that OTUs taxonomically
classified to \emph{Salmonella}, \emph{Escherichia},
\emph{Enterococcus}, and \emph{Staphylococcus} differed based on
polymerase used {[}Table S6 \& S7{]}. These specific OTUs were all
members of the mock community {[}Table S7{]}. The majority of the
differences between taxa were due to differences between Kappa and the
other polymerases {[}Table S7{]}. When differences in OTUs based on
polymerase were investigated for fecal samples, no differences could be
found (P-value \(\geqslant\) 0.01, corrected P-value = 1). These results
suggest that taxa specific differences in mock community samples do not
carry over to actual fecal community samples.

\textbf{\emph{Variation due to polymerase and number of cycles is
dwarfed by the biological variation of an individual.}} Using 16S rRNA
gene sequencing data there was a large disparity in a model's ability to
correctly classify samples based on polymerase used, number of cycles,
and individual {[}Figure 5A{]}. The model classification success for an
individual was significantly better than all other models tested
{[}Figure 5A \& Table S8{]}. Additionally, the top 10 most impotant
variables to the polymerase and number of cycle models were not the most
important variables in the model classifying individuals {[}Figure
5B{]}. This observation shows that OTUs that change due to the number of
cycles and polymerase used are not the same as OTUs that differentiate
between indvidiuals {[}Figure 5B{]}. Although there are a variety of
errors associated with the amplification process, they are smaller than
the bacterial community variation between indviduals.

### 2018-07-06 submission/supplemental.Rmd
add result updates

**Table S6: Dunns post-hoc results of differences in model classification success using 16S rRNA gene sequencing data**

@@

>>>>

**Table S6: Kruskal-Wallis results of significant differences in OTU in mock community samples**

@@

kruskal_otus <- read_csv("../data/process/tables/kruskal_otu_polymerase_mock_diffs_summary.csv") %>% 
  mutate(sample_type = "mock") %>% 
  filter(bh < 0.05) %>% 
  arrange(cycles, p.value) %>% 
  mutate_at(vars(p.value:bh),  funs(format(., scientific = TRUE, digits = 2, nsmall = 2))) %>% 
  mutate(genus = str_replace_all(genus, "_unclassified", ""))

colnames(kruskal_otus) <- c("Cycles", "OTU", "Lowest Taxa ID", "Statistic", "P-value", "BH", "Sample Type")

kable(kruskal_otus, format = "markdown", align = 'c', padding = 2)

**Table S7: Significant Dunns post-hoc results of differences between polymerases for specific taxa**

\footnotesize

```{r supplemental_table7, echo=FALSE, message=FALSE, warning=FALSE}

dunn_otus <- read_csv("../data/process/tables/dunns_sig_otu_polymerase_mock_diffs_summary.csv") %>% 
  arrange(bh) %>% 
  mutate_at(vars(chi2:bh),  funs(format(., scientific = TRUE, digits = 2, nsmall = 2)))

colnames(dunn_otus) <- c("Chi Sq", "Z", "P-value", "BH", "Comparison", "Cycles", "OTU", "Taxa", "Part of Mock")

kable(dunn_otus, format = "markdown", align = 'c', padding = 2)

\normalsize

**Table S8: Dunns post-hoc results of differences in model classification success using 16S rRNA gene sequencing data**

```{r supplemental_table8, echo=FALSE, message=FALSE, warning=FALSE}
